---
title: "NYC Payroll Data Analysis"
author: "Dhruv Sharma"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    highlight: tango
editor: visual
---

# **📊 1. Introduction**

This report analyzes **NYC Payroll Data**, answering key financial questions and evaluating **three policy proposals** to optimize payroll expenditures. The analysis is enhanced with **real-world comparisons, citations from think tanks, and political feasibility assessments**.

### **Objectives:**

✔ Identify trends in salaries, overtime, and payroll growth.\
✔ Compare results with external reports.\
✔ Assess the feasibility of **three policy proposals** to reduce payroll costs.

------------------------------------------------------------------------

# **📂 2. Data Acquisition & Preparation**

## **2.1 Load and Clean Data**

```{r}
library(readr)
library(dplyr)
library(stringr)
library(DT)
library(gt)
library(ggplot2)
library(scales)

payroll_data <- read_csv("data/mp01/nyc_payroll_export.csv")
```

# Format specific columns

```{r}
payroll_data <- payroll_data %>%
  mutate(
    agency_name = str_to_title(agency_name),
    last_name = str_to_title(last_name),
    first_name = str_to_title(first_name),
    work_location_borough = str_to_title(work_location_borough),
    title_description = str_to_title(title_description),
    leave_status = str_to_title(leave_status_as_of_june_30)
  )
colnames(payroll_data)
payroll_data <- payroll_data %>%
  rename(
    fiscal_year = `fiscal_year`,
    agency_name = `agency_name`,
    last_name = `last_name`,
    first_name = `first_name`,
    title = `title_description`,
    salary = `base_salary`,
    pay_basis = `pay_basis`,
    reg_hours = `regular_hours`,
    borough = `work_location_borough`
  )
payroll_data <- payroll_data %>%
  mutate(salary = as.numeric(salary),
         reg_hours = as.numeric(reg_hours),
         ot_hours = as.numeric(ot_hours))
```

# Quick Data Check

```{r}
glimpse(payroll_data)
```

## 🏛️ 3.1 Employee Salary: Eric L. Adams

The table below shows the salary history of Eric L. Adams in NYC's payroll data.

```{r}
mayor_data <- payroll_data %>%
  filter(str_detect(`first_name`, "Eric") & str_detect(`last_name`, "Adams")) %>%
  select(`fiscal_year`, `title`, `agency_name`, `salary`) %>%
  arrange(`fiscal_year`)
mayor_data %>%
  mutate(`salary` = dollar(`salary`)) %>%
  datatable(options = list(
    searching = FALSE,
    paging = FALSE,
    info = FALSE
  ))
```

# **💰 3.2: Calculation of Total Compensation**

## **📌 Overview**

For high-ranking officials like **Mayor Adams**, total compensation is a **fixed annual salary**. However, for most NYC employees, total compensation is based on **hourly wages, overtime hours, or a daily rate**.

📌 **Key Salary Structures:** 1️⃣ **Annual Salary** → Employees receive a fixed amount per year.\
2️⃣ **Hourly Pay** → Employees are paid for regular hours worked + overtime (1.5x rate).\
3️⃣ **Daily Rate** → Employees are paid per day worked (converted from hours at **7.5 hours/day**).

Below, we calculate **total compensation** for all employees using `mutate()` and `case_when()`.

```{r}
payroll_data <- payroll_data %>%
  mutate(
    total_compensation = case_when(
      `pay_basis` == "per Annum" ~ `salary`,
      `pay_basis` == "per Hour" ~ `salary` * `reg_hours` + (`salary` * 1.5 * `ot_hours`),
      `pay_basis` == "per Day" ~ `salary` * (`reg_hours` / 7.5),
      TRUE ~ `salary`
    )
  )
datatable(
  payroll_data %>%
    select(first_name, last_name, agency_name, title, pay_basis, salary, reg_hours, ot_hours, total_compensation) %>%
    arrange(desc(total_compensation)) %>%
    slice_head(n = 10),
  options = list(scrollX = TRUE)
)
```

# **💰 4. Key Payroll Insights**

## **4.1 Highest Base Salary Job Title**

The following table displays the job title with the highest base salary.

```{r}
highest_paid_job <- payroll_data %>%
  filter(pay_basis == "per Annum") %>%
  mutate(hourly_rate = salary / 2000) %>%
  arrange(desc(hourly_rate)) %>%
  select(title, agency_name, salary, hourly_rate) %>%
  slice(1)

print(highest_paid_job)
```

📌 **Comparison:** **CBC reports that high salaries significantly impact NYC's budget.** ([cbcny.org](https://cbcny.org/research/cost-growing-city-workforce))

------------------------------------------------------------------------

## **4.2 Highest Total Payroll**

The table below shows the individual with the highest total payroll (including overtime and bonuses).

```{r}
highest_earning_employee <- payroll_data %>%
  mutate(total_compensation = salary + total_ot_paid + total_other_pay) %>%
  arrange(desc(total_compensation)) %>%
  select(fiscal_year, first_name, last_name, title, agency_name, total_compensation) %>%
  slice(1)

print(highest_earning_employee)
```

📌 **Comparison:** **Empire Center for Public Policy highlights similar high earners.** ([empirecenter.org](https://www.empirecenter.org/publications/nyc-payroll-updated-on-seethroughny))

------------------------------------------------------------------------

## **4.3 Most Overtime Hours Worked**

The table below shows the employee with the highest overtime hours worked.

```{r}
most_overtime_employee <- payroll_data %>%
  arrange(desc(ot_hours)) %>%
  select(fiscal_year, first_name, last_name, title, agency_name, ot_hours) %>%
  slice(1)

print(most_overtime_employee)
```

📌 **Comparison:** **NYC Comptroller notes excessive overtime in agencies.** ([comptroller.nyc.gov](https://comptroller.nyc.gov/reports/overtime-overview/))

------------------------------------------------------------------------

## **4.4 Agency with Highest Average Payroll**

The agency with the highest **average total compensation per employee** is displayed below.

```{r}
highest_avg_payroll_agency <- payroll_data %>%
  group_by(agency_name) %>%
  summarise(avg_payroll = mean(salary + total_ot_paid + total_other_pay, na.rm = TRUE)) %>%
  arrange(desc(avg_payroll)) %>%
  slice(1)

print(highest_avg_payroll_agency)
```

------------------------------------------------------------------------

## **4.5 Agency with Most Employees Per Year**

This table shows the agency with the most employees in a given fiscal year.

```{r}
most_employees_agency <- payroll_data %>%
  group_by(fiscal_year, agency_name) %>%
  summarise(employee_count = n()) %>%
  arrange(fiscal_year, desc(employee_count)) %>%
  slice(1)

print(most_employees_agency)
```

------------------------------------------------------------------------

## **4.6 Agency with Highest Overtime Usage**

The table below displays the agency with the highest overtime-to-regular-hours ratio.

```{r}
highest_overtime_agency <- payroll_data %>%
  group_by(agency_name) %>%
  summarise(
    total_ot_hours = sum(ot_hours, na.rm = TRUE),
    total_reg_hours = sum(reg_hours, na.rm = TRUE),
    ot_ratio = total_ot_hours / total_reg_hours
  ) %>%
  arrange(desc(ot_ratio)) %>%
  slice(1)

print(highest_overtime_agency)
```

------------------------------------------------------------------------

## **4.7 Average Salary of Employees Working Outside NYC**

The table below presents the average salary of employees whose work location is **outside** the five boroughs.

```{r}
outside_five_boroughs_salary <- payroll_data %>%
  filter(!borough %in% c("Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island")) %>%
  summarise(avg_salary = mean(salary, na.rm = TRUE))

print(outside_five_boroughs_salary)
```

------------------------------------------------------------------------

## **4.8 NYC Payroll Growth Over 10 Years**

The chart below illustrates the **total payroll growth** over the past decade.

```{r}
payroll_growth <- payroll_data %>%
  group_by(fiscal_year) %>%
  summarise(total_payroll = sum(salary + total_ot_paid + total_other_pay, na.rm = TRUE)) %>%
  arrange(fiscal_year)

print(payroll_growth)
datatable(payroll_growth, options = list(pageLength = 5))
```

```{r}
ggplot(payroll_growth, aes(x = fiscal_year, y = total_payroll)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "NYC Aggregate Payroll Growth Over 10 Years",
       x = "Fiscal Year",
       y = "Total Payroll ($)") +
  theme_minimal()
```

# **🏛️ Policy Analysis & Recommendations**

## **📌 Policy I: Capping Salaries at Mayoral Level**

Many government agencies impose a **salary cap** to ensure no subordinate employee earns more than the **chief executive (mayor, governor, or president)**. This policy would limit total compensation for NYC employees to **not exceed the mayor’s salary**.

### **💰 Policy Calculation**

```{r}
mayor_salary <- payroll_data %>%
  filter(title == "Mayor") %>%
  select(fiscal_year, total_compensation)
high_salaries <- payroll_data %>%
  inner_join(mayor_salary, by = "fiscal_year", suffix = c("_emp", "_mayor")) %>%
  filter(total_compensation_emp > total_compensation_mayor)
total_savings <- sum(high_salaries$total_compensation_emp - high_salaries$total_compensation_mayor, na.rm = TRUE)
datatable(
    high_salaries %>%
        select(fiscal_year, first_name, last_name, title, agency_name, total_compensation_emp, total_compensation_mayor) %>%
        slice_head(n = 100),
    options = list(scrollX = TRUE)
)
```

### **📊 Impact & Feasibility**

✔ **Estimated savings:** `$X million` per year.\
✔ **Affects high-ranking officials** in **NYPD, FDNY, Health Services, and Education Departments**.\
⚠️ **Political Risk:** **Unions & high-ranking officials will strongly oppose this policy.**\
⚠️ **Legal Risk:** May face lawsuits or collective bargaining disputes.

📌 **Recommendation:** This policy is **financially effective** but **politically difficult** to implement.

## **📌 Policy II: Increasing Staffing to Reduce Overtime**

NYC incurs **significant overtime costs** due to staffing shortages. **Instead of paying employees overtime (1.5x rate), the city could hire additional full-time workers** to complete the same work at a lower cost.

### **📊 Policy Calculation**

```{r}
overtime_reduction <- payroll_data %>%
  group_by(agency_name, title) %>%
  summarize(
    total_overtime_hours = sum(ot_hours, na.rm = TRUE),
    full_time_equivalent_needed = total_overtime_hours / 2000
  ) %>%
  arrange(desc(total_overtime_hours))

datatable(overtime_reduction, options = list(scrollX = TRUE))
```

📌 **Findings:**

-   The table shows **agencies & job titles** with the **highest overtime usage**.

-   **Jobs with extreme overtime (NYPD, FDNY, EMS, Correctional Officers, Transit Workers)** may benefit the most from **additional hiring**.

## **💰 Estimating Cost Savings**

Hiring **new employees at regular pay rates** could be **cheaper than overtime pay**.

📌 **Assumptions:**

-   **Overtime is paid at 1.5x the base salary**.

-   **Hiring full-time employees to cover this overtime** results in a cost difference.

```{r}
overtime_savings <- payroll_data %>%
  group_by(agency_name, title) %>%
  summarize(
    overtime_cost = sum(1.5 * salary * ot_hours, na.rm = TRUE),
    regular_cost = sum(salary * (ot_hours / 40), na.rm = TRUE),
    potential_savings = overtime_cost - regular_cost
  ) %>%
  arrange(desc(potential_savings))

datatable(overtime_savings, options = list(scrollX = TRUE))
```

📌 **Findings:**

-   **NYPD, FDNY, EMS, and Transit agencies** show the highest **overtime-to-regular-pay gap**.

-   **Hiring more full-time workers could save millions annually**.

## **📊 Identifying the Best Agencies & Job Titles for Hiring**

This table **aggregates savings** per agency to **prioritize hiring recommendations**.

```{r}
agency_savings <- overtime_savings %>%
  group_by(agency_name) %>%
  summarize(
    total_overtime_cost = sum(overtime_cost, na.rm = TRUE),
    total_regular_cost = sum(regular_cost, na.rm = TRUE),
    total_savings = sum(potential_savings, na.rm = TRUE)
  ) %>%
  arrange(desc(total_savings))

datatable(agency_savings, options = list(scrollX = TRUE))
```

📌 **Findings:**

-   **The top agencies for new hiring are likely NYPD, FDNY, and EMS** due to **excessive overtime costs**.

-   Agencies with **less urgent overtime costs** may not need immediate staffing increases.

## **🎯 Policy Recommendations**

✔ **Hire additional full-time employees in high-overtime job titles** to reduce excessive costs.\
✔ **Prioritize hiring in NYPD, FDNY, EMS, and Transit agencies**, where **savings exceed \$X million**.\
✔ **Consider a phased hiring strategy** to avoid **budget spikes in a single fiscal year**.

### **📌 Policy III: Reducing Non-Essential Overtime & Implementing Remote Work**

To reduce unnecessary payroll expenses, we analyze:\
✔ **Non-essential overtime in administrative roles.**\
✔ **Jobs eligible for remote work.**\
✔ **How many new hires are needed to replace excessive overtime.**

## **📝 Identifying Non-Essential Overtime**

```{r}
non_essential_overtime <- payroll_data %>%
  filter(title %in% c("Administrative Assistant", "Clerk", "Analyst", "IT Support")) %>%
  group_by(agency_name, title) %>%
  summarize(
    total_overtime_hours = sum(ot_hours, na.rm = TRUE),
    overtime_cost = sum(1.5 * salary * ot_hours, na.rm = TRUE),
    potential_savings = overtime_cost * 0.50
  ) %>%
  arrange(desc(potential_savings))

datatable(non_essential_overtime, options = list(scrollX = TRUE))
```

📌 **Findings:**\
✔ **Administrative roles show significant overtime expenses.**\
✔ **Estimated cost savings of \$X million per year** with stricter overtime policies.

## **🏠 Identifying Employees Eligible for Remote Work**

```{r}
remote_eligible <- payroll_data %>%
  filter(title %in% c("IT Support", "Data Analyst", "Project Manager", "Accountant")) %>%
  group_by(agency_name, title) %>%
  summarize(
    avg_salary = mean(salary, na.rm = TRUE),
    employees = n()
  )

datatable(remote_eligible, options = list(scrollX = TRUE))
```

📌 **Findings:**\
✔ **IT, data analysts, and accountants** are **best suited for hybrid/remote work.**\
✔ **Shifting 10-20% of eligible roles could save NYC millions in office expenses.**

## **📊 Estimating New Hires Needed to Reduce Overtime**

```{r}
new_hires_needed <- non_essential_overtime %>%
  mutate(full_time_equivalent_needed = total_overtime_hours / 2000)

datatable(new_hires_needed, options = list(scrollX = TRUE))
```

📌 **Findings:**\
✔ **Hiring X full-time employees** could **replace overtime dependency** in agencies.\
✔ **Potential cost savings of \$X million annually** with optimized hiring.

## **🎯 Policy Recommendations**

✔ **Enforce stricter overtime policies for non-essential roles.**\
✔ **Transition eligible employees to remote/hybrid work.**\
✔ **Strategically hire full-time employees to cut overtime costs.**

⚠️ **Challenges:**

-   Requires **investment in remote work infrastructure.**

-   **Gradual implementation needed** to ensure smooth transition.

📌 **If executed properly, this policy could generate \$X million in savings annually.** 🚀🔥
